---
title: "EDA"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(dplyr)

```

#Introduction section

##Briefly describes the data and variables you have to work with

##Describe the purpose of your EDA and ultimate goal of modeling


#Data: Use a relative path to import data 


```{r,results="hide}

diabetes_data <- read.csv("diabetes_binary_health_indicators_BRFSS2015.csv")

diabetes_data <- as_tibble(diabetes_data)

diabetes_data <- diabetes_data |>
  drop_na() |>
  mutate(Diabetes_binary =factor(Diabetes_binary))

diabetes_data <- diabetes_data |>
  drop_na() |>
  mutate(across(c(Diabetes_binary, HighBP,
                  HighChol, CholCheck,
                  Smoker,Stroke),factor))

summary(diabetes_data)
str(diabetes_data)
```

#Summarizations: Univariate 
#Looking at summary statistics of each of the three variables
#Across the whole data set

```{r}

df_model <- diabetes_data |>
  drop_na() |>
  mutate(across(c(Diabetes_binary, HighChol,
                  PhysActivity, HighBP),factor)) |>
  select(Diabetes_binary, HighChol, BMI, PhysActivity,HighBP)

#Obtain summary statistics
summary(df_model)
str(df_model)


#Graph for BMI
ggplot(data=df_model, aes(x=BMI))+
  geom_histogram(binwidth=2)

#Graph for High Blood Pressure Distribution 
ggplot(data=df_model, aes(x=HighBP)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title ="High Blood Pressure Distribution", x="High Blood Pressure", y="Count")+
  scale_x_discrete(labels =c("No",
                             "Yes"))

#Graph for Physical Activity 
ggplot(data=df_model, aes(x=PhysActivity)) +
  geom_bar() +
  labs(title ="Physical Activity Distribution", x="Physical Activity", y="Count")+
  scale_x_discrete(labels =c("No",
                             "Yes"))
#Graph for Diabetes 
ggplot(data=df_model, aes(x=Diabetes_binary)) +
  geom_bar() +
  labs(title ="Diabetes Distribution", x="Diabetes", y="Count")+
  scale_x_discrete(labels =c("No",
                             "Yes"))

```
#Look at relationship between diabetes and each of the 
#indepedent variables

```{r}
#Diabetes + BMI

#Diabetes + Physical Activity

#Diabetes + High Blood Pressure

#Adding interactions between the three response variables



```
